An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-30 paper/main.tex
fertig

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{siunitx}

@@

\noindent
Another work is Q-BERT~\cite{shen2020qbert}, that focuses on transformer-based
language models. BERT (Bidirectional Encoder Representations from Transformers)
is a large transformer architecture pretrained on large text data and used as a
generic language representation model. Q-BERT shows that the weights of BERT
can be quantized down to 4 bits while maintaining accuracy close to the
full-precision model.

@@

\noindent
In our MLP, hidden layers use the Rectified Linear Unit (ReLU), so
$\sigma=\mathrm{ReLU}$, while the output layer is linear, producing raw scores
(also called logits) $z$ later used in~\eqref{eq:ce}. ReLU~\eqref{eq:relu}
keeps positive activations linear and sets negatives to zero. An additional
benefit is that the calculation is very fast and simple compared to more
complex activation functions.

@@

\begin{equation}\label{eq:cegrads}
	\frac{\partial L}{\partial z_k}=p_k-\mathbbm{1}\{k=y\},\qquad\frac{\partial L}{\partial W}=(p-e_y)h^\top,\qquad\frac{\partial L}{\partial b}=p-e_y,
\end{equation}

with $W$ and $b$ the weight matrix and bias of the last linear layer, $h$ the
layer input, and $e_y$ the one-hot vector of the target class. Here,
$\mathbbm{1}\{k=y\}$ is the indicator: it equals \num{1} if $k=y$ and \num{0}
otherwise.\\

@@

\begin{table}[H]
	\centering
	\begin{tabular}{@{}lcccc@{}}
		\toprule
		\textbf{Type} & Sign & Exponent & Significand & Total \\
		\midrule
		Half  (FP16)  & 1    & 5        & 10          & 16    \\
		Single (FP32) & 1    & 8        & 23          & 32    \\
		Double (FP64) & 1    & 11       & 52          & 64    \\
		\bottomrule
	\end{tabular}
	\caption{Field widths (in bits) of common IEEE754 formats (adapted from~\cite{fparithmetic}).}\label{tab:binary-bits}
\end{table}

\begin{table}[H]
	\centering
	\begin{tabular}{@{}lccc@{}}
		\toprule
		\textbf{Type} & Exponent bias & Precision (bits) & Decimal digits \\
		\midrule
		Half (FP16)   & 15            & 11               & $\sim$3.3      \\
		Single (FP32) & 127           & 24               & $\sim$7.2      \\
		Double (FP64) & 1023          & 53               & $\sim$15.9     \\
		\bottomrule
	\end{tabular}
	\caption{Range and precision characteristics of the formats (adapted from~\cite{fparithmetic}).}\label{tab:binary-precision}
\end{table}

@@

\section{Experiments and Results}\label{sec:results}
In the first experiment, every floating-point tensor that is used inside the
training loop was converted to a lower precision data type. This therefore
includes the weights and biases, activations, gradients, optimizer states, etc.
The types tried were FP64, FP32, and FP16.

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1\%$. Retraining the
models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the FP64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with different precisions.}\label{fig:modelAccTrainAndInf}
\end{figure}

\subsection*{Training and Inference Time}
We also analyzed the times needed for both, training and inference.
Figure~\ref{fig:modelEvalTrainTimes} depicts the times measured in the training
loop and later on at the inference stage. Again, the times are almost identical
in both regards, and neither a real speed-up nor a slow-down has been measured.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/times.png}
	\caption{Training and evaluation times for the different precisions.}\label{fig:modelEvalTrainTimes}
\end{figure}

@@

\subsection*{Automatic Mixed Precision and Gradient Clipping}
AMP improves deep learning training by using both FP16 and FP32 data types
reducing memory usage and increasing computational speed. However, FP16's
limited dynamic range can cause numerical instability, especially during
backpropagation. To address this, Gradient Clipping is used to cap the
magnitude of gradients (either globally or per element) preventing extreme
updates that could hinder convergence. Together, AMP and gradient clipping
enable efficient, stable training by balancing performance and numerical
precision.

First step is using global norm clipping, all gradients are collectively scaled
to ensure their combined norm does not exceed a specified threshold, promoting
stable updates across the network. In second step per-element clipping
constrains each individual gradient component within a fixed range, directly
limiting extreme values. These methods are evaluated under AMP to compare their
impact on training stability, model accuracy, and computation time across
different precision types.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{figures/FP16_32_clipping.png}
	\caption{Affect of Gradient Norm on accuracy, training and testing timings.
		Accuracy is virtually the same between FP16 and FP32 when gradient clipping is used
		validating the stability of AMP and gradient scaling.
		Eval time is nearly identical for both formats, as evaluation involves fewer operations
		and less intense memory bandwidth usage.}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{figures/gradclip_16.png}
	\caption{Effect of Per-Element Gradient Clipping on FP16. Accuracy vs Training \& Eval Time}\label{fig:grad16}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{figures/gradclip_32.png}
	\caption{Effect of Per-Element Gradient Clipping on FP32. Accuracy vs Training \& Eval Time}\label{fig:grad32}
\end{figure}

Figure~\ref{fig:grad16} shows accuracy peaked at 97.07\% with a clipping value
of 0.5, while the lowest accuracy (96.55\%) occurred at clip=0.01, suggesting
that very small clip values may overly restrict gradient updates and hurt
learning. Training time ranged indicating that moderate clipping supports
efficient convergence. Evaluation time remained stable across all clip values
(around 2.01s to 2.16s), showing that clipping primarily impacts training
rather than inference performance. Also Figure~\ref{fig:grad32} shows that
accuracy was most stable across all clip values with the highest accuracy at
clip=0.5, demonstrating its robustness to gradient clipping variations.
Training time suggesting that larger clip values can speed up convergence
without sacrificing stability. Evaluation time confirming that clipping had
minimal effect on inference time.

\subsection*{Effect of Decimal Rounding}

A non training related issue is the storage of the NN parameters. Usually, the
parameters are stored in the format that was chosen to train with. This is best
for reproducibility as well as accuracy when doing inference with the trained
model weights. The datatype precision has a higher impact during training
though, which begs the question if one could omit precision from the model
weights. In this section, we explore a naive quantization method: Rounding to a
fixed number of decimals. With significantly little decimals and a reasonable
range for the integer part, one could store only the significant and provide
base as well as exponent as metadata. When loading the model, the weights are
converted back to floating point, with whatever precision is desired. Only the
weights are quantized; the input data, as well are the values propagated
through the network are at full precision.

The first step is to examine the distribution of parameter values. By the
standard deviation around modes one can make an educated guess as to how many
decimals are needed. The range will decide how large the integer part may get.
Reading from the distribution is just a heuristic though, so we will also test
the hypotheses.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_dist.png}
	\caption{
		Distribution of model weights in each layer. The values are in $[-0.5, 0.5]$,
		this means no integer part is needed in this example. Generally, with normalized data
		and the quite popular L1-/L2-/Batch-Norm layers, weight values in $[-1, 1]$ is to be expected.}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_dist2.png}
	\caption{Distribution of model weights in each layer after rounding to two decimals.
		This appears to be a good approximation already.}\label{fig:118_2}
\end{figure}

@@

Figure~\ref{fig:118_acc} shows how rounding the decimal part affects the test
accuracy. For just one decimal the model has a maximum of 19 different weights
at its disposal, which makes the model have only 10\% accuracy, random guessing
with a 10 class problem. Two decimals perform significantly better. At 70\%
accuracy the model gets close to the full precision 88\% accuracy with 32 bit
floating point. Making an educated guess on distribution such as
Figure~\ref{fig:118_2} seems to be a good heuristic for this example. At just 3
decimals, the full 32 bit precision is reached. Thus we can conclude that with
just 10 bit for representing 3 decimals, no bits for the integer part and one
additional bit for the sign, only 11 bit must be stored to have the best
performing model for inference.

\section{Summary}

The experimental results presented in this paper demonstrate the robustness of
neural networks to reduced parameter precision, particularly for simpler tasks
like MNIST classification. Key findings include:

\begin{itemize}
	\item \textbf{Precision Reduction During Training and Inference}: Training and
	      evaluating models with 16-bit floating-point (FP16) precision preserved accuracy
	      comparable to higher-precision formats (FP32 and FP64). This suggests that for
	      tasks of similar complexity, FP16 is sufficient, offering potential memory and
	      computational savings without sacrificing performance.

	\item \textbf{Post-Training Quantization}: Reducing precision only during inference, after
	      training with high-precision (FP64), maintained baseline accuracy. This is
	      particularly valuable for deployment scenarios where training is done on powerful
	      hardware, but inference occurs on resource-constrained devices.

	\item \textbf{Mixed-Precision:} Training with mixed precision and Gradient Clipping
	      resulted in similar test accuracy scores as precision reduction during training
	      as well as post-training quantization. No significant decrease in model
	      performance accuracy was detected. The train and test times were similar,
	      regardless of using FP32, FP16 or a mix of both.

	\item \textbf{Weight Rounding:} Rounding trained weights to just three decimal digits
	      (approximately 11 bits) achieved full accuracy, highlighting the potential for
	      aggressive quantization in storage-efficient model deployment. This finding is
	      significant for edge applications where memory footprint is critical.
\end{itemize}

However, the simplicity of the MNIST dataset and the small network architecture
used in the experiments may limit the generalizability of these findings. More
complex tasks (e.g., CIFAR-10, ImageNet) and deeper architectures (e.g., CNNs,
Transformers) could exhibit greater sensitivity to precision reduction.
Additionally, the impact of precision on training dynamics, such as gradient
flow and convergence, was not deeply explored and could be a focus for future
work.

\section{Conclusion}

This study demonstrates that neural networks can maintain strong performance
even under significant reductions in parameter precision, particularly for
simpler tasks like MNIST classification. Key findings confirm that FP16
training and inference, post-training quantization, and aggressive weight
rounding (down to three decimal digits) preserve model accuracy while improving
memory and computational efficiency. These results are promising for deploying
models in resource-constrained environments, such as edge devices and real-time
applications.

To build on these insights and expand their applicability, future work should
focus on the following directions:

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsmath, amssymb}
\usepackage{graphicx}
\usepackage[hidelinks]{hyperref}
\usepackage{geometry}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{csquotes}
\usepackage{float}
\usepackage{derivative}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{siunitx}
\usepackage{refcount}

@@

\noindent
Another work is Q-BERT~\cite{shen2020qbert}, that focuses on transformer-based
language models. BERT (Bidirectional Encoder Representations from Transformers)
is a large transformer architecture pretrained on large text data and used as a
generic language representation model. Q-BERT shows that the weights of BERT
can be quantized down to 4 bits while maintaining accuracy close to the
full-precision model.\\

\noindent
AMP improves deep learning training by using both FP16 and FP32 data types
reducing memory usage and increasing computational speed. However, FP16's
limited dynamic range can cause numerical instability, especially during
backpropagation. To address this, Gradient Clipping is used to cap the
magnitude of gradients (either globally or per element) preventing extreme
updates that could hinder convergence. Together, AMP and gradient clipping
enable efficient, stable training by balancing performance and numerical
precision~\cite{micikevicius2017mixed}.

@@

\noindent
In our MLP, hidden layers use the Rectified Linear Unit (ReLU), so
$\sigma:=\mathrm{ReLU}$, while the output layer is linear, producing raw scores
(also called logits) $z$ later used in~\eqref{eq:ce}. ReLU~\eqref{eq:relu}
keeps positive activations linear and sets negatives to zero. An additional
benefit is that the calculation is very fast and simple compared to more
complex activation functions.

@@

\begin{equation}\label{eq:cegrads}
	\frac{\partial L}{\partial z_k}=p_k-\mathbbm{1}_{k=y},\qquad\frac{\partial L}{\partial W}=(p-e_y)h^\top,\qquad\frac{\partial L}{\partial b}=p-e_y,
\end{equation}

with $W$ and $b$ the weight matrix and bias of the last linear layer, $h$ the
layer input, and $e_y$ the one-hot vector of the target class. Here,
$\mathbbm{1}_{k=y}$ is the indicator: it equals \num{1} if $k=y$ and \num{0}
otherwise. For evaluation, the model is fed multiple inputs and the results of
the indicator function are summed up and divided by the total number of inputs.
This is then called the accuracy score of the model which can be understood as
the ratio of correct predictions.\\

@@

\begin{table}[H]
	\centering
	\begin{tabular}{@{}lcccc@{}}
		\toprule
		\textbf{Type} & Sign & Exponent & Significand & Total \\
		\midrule
		Half  (FP16)  & 1    & 5        & 10          & 16    \\
		Single (FP32) & 1    & 8        & 23          & 32    \\
		Double (FP64) & 1    & 11       & 52          & 64    \\
		\bottomrule
	\end{tabular}
	\caption{Field widths (in bits) of common IEEE754 formats\protect\footnotemark.}\label{tab:binary-bits}
\end{table}
\footnotetext{\label{footnoteWiki}\url{https://en.wikipedia.org/wiki/Floating-point_arithmetic}}

\begin{table}[H]
	\centering
	\begin{tabular}{@{}lccc@{}}
		\toprule
		\textbf{Type} & Exponent bias & Precision (bits) & Decimal digits \\
		\midrule
		Half (FP16)   & 15            & 11               & $\sim$3.3      \\
		Single (FP32) & 127           & 24               & $\sim$7.2      \\
		Double (FP64) & 1023          & 53               & $\sim$15.9     \\
		\bottomrule
	\end{tabular}
	\caption{Range and precision characteristics of the formats\protect\footnotemark[\getrefnumber{footnoteWiki}].}\label{tab:binary-precision}
\end{table}

@@

\section{Experiments and Results}\label{sec:results}
In the first experiment, every floating-point variable that is used inside the
training loop was converted to a lower precision data type. This therefore
includes the weights and biases, activations, gradients, and optimizer states.
The types tried were FP64, FP32, and FP16.

\subsection*{Accuracy Across Bitwidths}
Figure~\ref{fig:modelAccTrainAndInf} shows how the accuracy changed after the
models were trained on the different precision settings. Keep in mind that the
y-axis is not starting from zero, so the absolute height differences are
amplified visually, though the delta is only around $.1\%$. Retraining the
models and re-evaluating them has also lead to the accuracies being
``reversed'', meaning the FP64 model had the least accuracy. It is safe to
assume that the model accuracy has neither decreased, nor increased.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/trainingAndInference.png}
	\caption{Model accuracy evaluated after training with different precisions (5-fold CV).}\label{fig:modelAccTrainAndInf}
\end{figure}

\subsection*{Training and Inference Time}
We also analyzed the times needed for both, training and inference.
Figure~\ref{fig:modelEvalTrainTimes} depicts the times measured in the training
loop and later on at the inference stage. Again, the times are almost identical
in both regards, and neither a real speed-up nor a slow-down has been measured.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/times.png}
	\caption{Training and evaluation times for the different precisions (5-fold CV).}\label{fig:modelEvalTrainTimes}
\end{figure}

@@

\subsection*{Automatic Mixed Precision and Gradient Clipping}
This experiment uses AMP with different gradient clipping methods. The first
method is called global norm clipping where all gradients are collectively
scaled to ensure their combined norm does not exceed a specified threshold,
which ensures stable updates across the network. The second method is
per-element clipping which constrains each individual gradient component within
a fixed range, directly limiting extreme values. These methods are evaluated
under AMP to compare their impact on training stability, model accuracy, and
computation time across different precision types.\\

\noindent
Figure~\ref{fig:ampGradientNormClipping} depicts the effect of the
\textit{global} gradient norm on accuracy, training and testing timings.
Accuracy is virtually the same between FP16 and FP32 when gradient clipping is
used validating the stability of AMP and gradient scaling. Eval time is nearly
identical for both formats, as evaluation involves fewer operations and less
intense memory bandwidth usage.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{figures/FP16_32_clipping.png}
	\caption{Effect of global gradient norm clipping on accuracy, training and testing timings.}\label{fig:ampGradientNormClipping}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{figures/gradclip_16.png}
	\caption{Effect of per-element gradient clipping on FP16 $\rightarrow$ Accuracy vs training and evaluation time}\label{fig:grad16}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.95\textwidth]{figures/gradclip_32.png}
	\caption{Effect of per-element gradient clipping on FP32 $\rightarrow$ Accuracy vs training and evaluation time}\label{fig:grad32}
\end{figure}

\noindent
Figure~\ref{fig:grad16} shows that accuracy peaked at 97.07\% with a clipping
value of 0.5, while the lowest accuracy (96.55\%) occurred at clip=0.01,
suggesting that very small clip values may overly restrict gradient updates and
hurt learning. Training time varied, which indicates that moderate clipping
supports efficient convergence. The evaluation time remained stable across all
clip values (around 2.01s to 2.16s), showing that clipping primarily impacts
training rather than inference performance. Figure~\ref{fig:grad32} shows that
accuracy was most stable across all clip values with the highest accuracy at
clip=0.5, demonstrating its robustness to gradient clipping variations. The
training was conducted with 5-fold CV, 10 epochs, and batch size 64. The
training time suggests that larger clip values can speed up convergence without
sacrificing stability. Evaluation time confirms that clipping had only a minor
effect on inference time.

\subsection*{Effect of Decimal Rounding}
A non training related issue is the storage of the NN parameters. Usually, the
parameters are stored in the format that was chosen to train with. This is best
for reproducibility as well as accuracy when doing inference with the trained
model weights. The datatype precision has a higher impact during training
though, which begs the question if one could reduce the precision of the model
weights. In this section, we explore a naive quantization method: Rounding to a
fixed number of decimals. With a small enough number of decimals and a
reasonable range for the integer part, one could store only the significant and
provide base as well as exponent as separate data. When loading the model, the
weights are converted back to floating point, with whatever precision is
desired. Only the weights are quantized; the input data, as well are the values
propagated through the network are at full precision.\\

\noindent
The first step is to examine the distribution of parameter values. By the
standard deviation around modes one can make an educated guess as to how many
decimals are needed. The range will decide how large the integer part may get.
Reading from the distribution is just a heuristic though, so we will also test
the hypotheses.\\

\noindent
Figure~\ref{fig:roundingDistribution1} shows the distribution of the model
weights in a three layer architecture. The values are in $[-0.5, 0.5]$, which
means no integer part is needed in this example. In Figure~\ref{fig:118_2} the
weights of the same model rounded to two decimals is depicted. The envelope of
the bins is still very similar to the full precision distribution, which makes
it appear like a good approximation. The results of evaluating this heuristic
is shown in Figure 11, where the results of how rounding the decimal part
affects the test accuracy is shown.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_dist.png}
	\caption{
		Distribution of model weights in each layer.}\label{fig:roundingDistribution1}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_dist2.png}
	\caption{Distribution of model weights in each layer after rounding to two decimals.}\label{fig:118_2}
\end{figure}

@@

\noindent
Figure~\ref{fig:118_acc} shows how rounding the decimal part affects the test
accuracy. For just one decimal the model has a maximum of 19 different weights
at its disposal, which makes the model have only 10\% accuracy, random guessing
with a 10 class problem. Two decimals perform significantly better. At 70\%
accuracy the model gets close to the full precision 88\% accuracy with 32 Bit
floating point. Making an educated guess the on distribution such as
Figure~\ref{fig:118_2} seems to be a good heuristic for this example. At just 3
decimals, the full 32 bit precision is reached. Thus we can conclude that with
just \num{10} Bits for representing 3 decimals, no bits for the integer part
and one additional bit for the sign, only \num{11} Bits must be stored to have
the best performing model for inference. The storage size of the model is
reduced to $\frac{11}{32}\approx 34\%$ of the otherwise full \num{32} Bit
model.

\section{Summary and Conclusion}

Regarding the precision reduction during training and inference, models with
16-bit floating-point (FP16) precision preserved accuracy comparable to
higher-precision formats (FP32 and FP64). This suggests that for tasks of
similar complexity, FP16 is sufficient, offering potential memory and
computational savings without sacrificing performance. Reducing precision only
during inference, after training with high-precision (FP64), maintained
baseline accuracy. This is particularly valuable for deployment scenarios where
training is done on powerful hardware, but inference occurs on
resource-constrained devices. Training with mixed precision and Gradient
Clipping resulted in similar test accuracy scores as precision reduction during
training as well as post-training quantization. No significant decrease in
model performance accuracy was detected. The train and test times were similar,
regardless of using FP32, FP16 or a mix of both. Rounding trained weights to
just three decimal digits (approximately 11 bits) achieved full accuracy,
highlighting the potential for aggressive quantization in storage-efficient
model deployment. This finding is significant for edge applications where
memory footprint is critical.\\

\noindent
However, the simplicity of the MNIST dataset and the small network architecture
used in the experiments may limit the generalizability of these findings. More
complex tasks (e.g., CIFAR-10, ImageNet) and deeper architectures (e.g., CNNs,
Transformers) could exhibit greater sensitivity to precision reduction.
Additionally, the impact of precision on training dynamics, such as gradient
flow and convergence, was not deeply explored and could be a focus for future
work.\\

\noindent
This study demonstrates that neural networks can maintain strong performance
even under significant reductions in parameter precision, particularly for
simpler tasks like MNIST classification. Key findings confirm that FP16
training and inference, post-training quantization, and aggressive weight
rounding (down to three decimal digits) preserve model accuracy while improving
memory and computational efficiency. These results are promising for deploying
models in resource-constrained environments, such as edge devices and real-time
applications.\\

\noindent
To build on these insights and expand their applicability, future work should
focus on the following directions:
